---
title: "01_DESeq2"
output: html_document
date: "2026-03-25"
---

## 0. Configuration directories and libraries

Include all necessary libraries and paths to the corresponding files


```{r setup}

library(DESeq2)
library(edgeR)
library(limma)
library(ggplot2)
library(dplyr)
library(tibble)
library(RColorBrewer)
library(ggrepel)

in_dir   <- "/home/jose/Desktop/RNAseq/results/counts/bulk/"
diagnosis_dir <- "/home/jose/Desktop/RNAseq/data/Diagnosis_data"
out_dir  <- "/home/jose/Desktop/RNAseq/results/counts/bulk/DEG/DESeq2"
plot_dir <- "/home/jose/Desktop/RNAseq/results/plots/DESeq2"
dir.create(out_dir,  recursive = TRUE, showWarnings = FALSE)
dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)

in_counts   <- file.path(in_dir, "rawcounts_ComBatSeq.csv")
in_metadata <- file.path(diagnosis_dir, "Mostres_RNAseqJose.csv")

```

## 1. Thresholds

```{r thresholds}
FDR <- 0.05
FC <- 1
```

## 2. Load counts

```{r loading counts}

counts_df <- read.csv(in_counts, header = TRUE, stringsAsFactors = FALSE, check.names = FALSE)
colnames(counts_df)[1] <- "Gene"
counts_df[, -1] <- lapply(counts_df[, -1, drop = FALSE], as.numeric)

counts <- as.matrix(counts_df[, -1, drop = FALSE])
rownames(counts) <- counts_df$Gene
counts <- round(counts)

message("Counts loaded: ", nrow(counts), " genes x ", ncol(counts), " samples")

```

## 3. Build coldata

```{r coldata}

sample_names <- colnames(counts)

batch     <- ifelse(grepl("_FC1_", sample_names, ignore.case = TRUE), "FC1",
             ifelse(grepl("_FC2_", sample_names, ignore.case = TRUE), "FC2", "FC3"))
treatment <- ifelse(grepl("basal", sample_names, ignore.case = TRUE), "Basal", "LPS")
donor_id  <- sub("^[^_]+_[^_]+_([^_]+)_.*", "\\1", sample_names)
donor_id  <- gsub("\\.?LPS$", "", donor_id)

coldata <- data.frame(
  sample    = sample_names,
  treatment = factor(treatment, levels = c("Basal", "LPS")),
  batch     = factor(batch),
  donor_id  = factor(donor_id),
  stringsAsFactors = FALSE
)
rownames(coldata) <- sample_names

# Merge Diagnosis
meta_df       <- read.csv(in_metadata, header = TRUE, stringsAsFactors = FALSE, check.names = FALSE)
meta_id_col   <- "Sample_ID"   
meta_diag_col <- "Diagnosis"   

coldata$diagnosis <- meta_df[[meta_diag_col]][match(coldata$donor_id, meta_df[[meta_id_col]])]
coldata$diagnosis <- sub("\\+.*", "", coldata$diagnosis)  # keep first diagnosis

# Check all samples present a diagnosis
na_idx <- which(is.na(coldata$diagnosis))
if (length(na_idx) > 0) {
  stop("Missing diagnosis in samples:\n",
       paste(rownames(coldata)[na_idx], collapse = ", "))
} else {
  message("Diagnosis assigned to all samples successfully.")
}

# In case of lack of diagnosis may apply: "coldata$diagnosis[c(135,136,137,326,327,328)] <- "Tourette""

coldata$diagnosis <- factor(coldata$diagnosis)
message("Samples per group:")
print(table(coldata$treatment, coldata$diagnosis))


```

## 4. Gene Filtering

This step is made to filter genes that present few counts, that are not statistically quantifiable

```{r filter}

dge    <- DGEList(counts = counts, group = coldata$treatment)
keep   <- filterByExpr(dge, group = coldata$treatment)
counts <- counts[keep, ]
message("Genes after filtering: ", sum(keep), " / ", length(keep))

```

## 5. Plots gene regulation

```{r plots}

save_deg_results <- function(res, name, n_labels =15) {

  res_df <- as.data.frame(res) %>%
    rownames_to_column("gene") %>%
    arrange(padj)

  res_df$status <- "NS"
  res_df$status[res_df$padj < FDR & res_df$log2FoldChange >  FC] <- "Up"
  res_df$status[res_df$padj < FDR & res_df$log2FoldChange < -FC] <- "Down"
  res_df$status <- factor(res_df$status, levels = c("Up", "Down", "NS"))

  n_up   <- sum(res_df$status == "Up",   na.rm = TRUE)
  n_down <- sum(res_df$status == "Down", na.rm = TRUE)
  message(name, " — Up: ", n_up, "  Down: ", n_down)

  write.csv(res_df, file = file.path(out_dir, paste0(name, "_results.csv")), row.names = FALSE)

  res_filt <- res_df %>% filter(!is.na(padj))

  # ── Select genes to label ────────────────────────────────────────────────
  # Top N by significance among Up and Down genes, split evenly
  top_up <- res_filt %>%
    filter(status == "Up") %>%
    arrange(padj) %>%
    slice_head(n = ceiling(n_labels / 2))

  top_down <- res_filt %>%
    filter(status == "Down") %>%
    arrange(padj) %>%
    slice_head(n = floor(n_labels / 2))

  genes_to_label <- bind_rows(top_up, top_down)
  
  
  # Volcano
  p_vol <- ggplot(res_filt, aes(x = log2FoldChange, y = -log10(padj), color = status)) +
    geom_point(size = 1.2, alpha = 0.6) +
    scale_color_manual(values = c("Up"="#ee2c2c","Down"="#00008b","NS"="grey70")) +
    geom_vline(xintercept = c(-FC, FC), linetype="dashed", color="grey40") +
    geom_hline(yintercept = -log10(FDR),    linetype="dashed", color="grey40") +
    geom_text_repel(
      data          = genes_to_label,
      aes(label     = gene),
      size          = 3,
      color         = "black",
      max.overlaps  = 20,
      box.padding   = 0.4,
      segment.size  = 0.3,
      segment.color = "grey50",
      min.segment.length = 0
    ) +
    annotate("text", x =  max(res_filt$log2FoldChange)*0.8,
             y = max(-log10(res_filt$padj))*0.95,
             label = paste0("Up: ", n_up),   color="#ee2c2c", size=4) +
    annotate("text", x =  min(res_filt$log2FoldChange)*0.8,
             y = max(-log10(res_filt$padj))*0.95,
             label = paste0("Down: ", n_down), color="#00008b", size=4) +
    labs(title    = paste("Volcano —", name),
         subtitle = paste0("FDR < ", FDR, "  |  |log2FC| > ", FC,
                           "  |  Top ", nrow(genes_to_label), " genes labeled"),
         x = "log2 Fold Change", y = "-log10(FDR)") +
    theme_bw(base_size = 13) +
    theme(plot.title = element_text(face="bold"), legend.title = element_blank())

  ggsave(file.path(plot_dir, paste0(name, "_volcano.png")),
         p_vol, width=7, height=6, dpi=300)

  # MA plot
  p_ma <- ggplot(res_filt, aes(x = log2(baseMean+1), y = log2FoldChange, color = status)) +
    geom_point(size=1, alpha=0.5) +
    scale_color_manual(values = c("Up"="#ee2c2c","Down"="#00008b","NS"="grey70")) +
    geom_hline(yintercept = c(-FC, 0, FC),
               linetype=c("dashed","solid","dashed"), color="grey40") +
    labs(title    = paste("MA plot —", name),
         subtitle = paste0("FDR < ", FDR, "  |  |log2FC| > ", FC),
         x = "log2(mean expression)", y = "log2 Fold Change") +
    theme_bw(base_size=13) +
    theme(plot.title=element_text(face="bold"), legend.title=element_blank())

  ggsave(file.path(plot_dir, paste0(name, "_MA.png")),
         p_ma, width=7, height=6, dpi=300)

  invisible(res_df)
}

save_deg_results_lrt <- function(res, name) {

  res_df <- as.data.frame(res) %>%
    rownames_to_column("gene") %>%
    arrange(padj)

  res_filt <- res_df %>% filter(!is.na(padj))
  n_sig <- sum(res_filt$padj < FDR)
  message(name, " — Significant genes (FDR < ", FDR, "): ", n_sig)

  write.csv(res_df, file = file.path(out_dir, paste0(name, "_results.csv")),
            row.names = FALSE)

  # Plot — significancia vs expresión media
  p <- ggplot(res_filt, aes(x = log2(baseMean + 1), y = -log10(padj),
                            color = padj < FDR)) +
    geom_point(size = 1, alpha = 0.5) +
    scale_color_manual(values = c("TRUE" = "#ee2c2c", "FALSE" = "grey70"),
                       labels = c("TRUE" = paste0("FDR < ", FDR), "FALSE" = "NS")) +
    geom_hline(yintercept = -log10(FDR), linetype = "dashed", color = "grey40") +
    labs(title    = paste("LRT —", name),
         subtitle = paste0("Significant genes: ", n_sig),
         x = "log2(mean expression)", y = "-log10(FDR)", color = "") +
    theme_bw(base_size = 13) +
    theme(plot.title = element_text(face = "bold"))

  ggsave(file.path(plot_dir, paste0(name, "_LRT.png")),
         p, width = 7, height = 6, dpi = 300)

  invisible(res_df)
}

```

## 6. Treatment global model 

~ donor_id + treatment 
Comparisons LPS vs Basal in global

```{r treatment}

dds_treat <- DESeqDataSetFromMatrix(countData = counts, colData   = coldata, design = ~ donor_id + treatment)
dds_treat <- DESeq(dds_treat)
message("Treatment model coefficients:")
print(resultsNames(dds_treat))

# Comparison LPS vs Basal

res_A <- results(dds_treat, name  = "treatment_LPS_vs_Basal", alpha = FDR)
res_A <- lfcShrink(dds_treat, coef = "treatment_LPS_vs_Basal", type = "apeglm", res = res_A)
save_deg_results(res_A, "M1.1_LPS_vs_Basal")

```

## 7. Diagnosis differences in Basal vs LPS

~ donor_id + treatment (diagnosis subsets)
Comparisons LPS vs Basal in each diagnosis

```{r diagnosis}

diagnoses <- levels(coldata$diagnosis)

for (diag in diagnoses) {

  # Subset by diagnosis
 
  idx_diag <- coldata$diagnosis == diag
  counts_sub  <- counts[, idx_diag]
  coldata_sub <- coldata[idx_diag, ]

  # Generate coldata for each subset with clean factors
  coldata_sub$donor_id  <- droplevels(coldata_sub$donor_id)
  coldata_sub$treatment <- droplevels(coldata_sub$treatment)

  # Keep only PAIRED donors
 
  valid_donors <- coldata_sub %>%
    group_by(donor_id) %>%
    summarise(n_cond = n_distinct(treatment)) %>%
    filter(n_cond == 2) %>%
    pull(donor_id)

  # Subset to paired donors
  idx_paired <- coldata_sub$donor_id %in% valid_donors

  counts_sub  <- counts_sub[, idx_paired]
  coldata_sub <- coldata_sub[idx_paired, ]

  # Clean again after subsetting
  coldata_sub$donor_id  <- droplevels(coldata_sub$donor_id)
  coldata_sub$treatment <- droplevels(coldata_sub$treatment)

  # Safety checks
 
  if (ncol(counts_sub) == 0) {message("Skipping ", diag, " (no paired samples)")
    next}

  if (length(unique(coldata_sub$treatment)) < 2) {message("Skipping ", diag, " (only one treatment after filtering)")
    next}

  if (length(unique(coldata_sub$donor_id)) < 2) {message("Skipping ", diag, " (not enough donors)")
    next}

  message("Samples: ", ncol(counts_sub),
          " | Donors: ", length(unique(coldata_sub$donor_id)))

  print(table(coldata_sub$donor_id, coldata_sub$treatment))

  # DESeq2 model (paired)
  dds_sub <- DESeqDataSetFromMatrix(countData = counts_sub, colData = coldata_sub, design = ~ donor_id + treatment)
  dds_sub <- DESeq(dds_sub)

  message("Coefficients:")
  print(resultsNames(dds_sub))
  
  res <- results(dds_sub, name  = "treatment_LPS_vs_Basal", alpha = FDR)
  res <- lfcShrink(dds_sub, coef = "treatment_LPS_vs_Basal", type = "apeglm", res  = res)

  save_deg_results(res, paste0("M2_LPS_vs_Basal_", diag))
}
```

## 8. Diagnosis differences Basal

~ diagnosis in Basal subset
Comparisons global + pairwise (ASD-OCD, ASD-TS, OCD-TS) without LPS noise

```{r diagnosis}

idx_basal     <- coldata$treatment == "Basal"
counts_basal  <- counts[, idx_basal]
coldata_basal <- coldata[idx_basal, ]
coldata_basal$diagnosis <- droplevels(coldata_basal$diagnosis)

dds_basal <- DESeqDataSetFromMatrix(countData = counts_basal, colData = coldata_basal, design = ~ diagnosis)
dds_basal <- DESeq(dds_basal)

# ASD vs OCD

res_Bx1 <- results(dds_basal, contrast = c("diagnosis", "TEA", "TOC"), alpha = FDR)
save_deg_results(res_Bx1, "M3.1_Basal_ASD_vs_OCD")

# ASD vs TS

res_Bx2 <- results(dds_basal, contrast = c("diagnosis", "TEA", "Tourette"), alpha = FDR)
save_deg_results(res_Bx2, "M3.2_Basal_ASD_vs_TS")

# OCD vs TS

res_Bx3 <- results(dds_basal, contrast = c("diagnosis", "TOC", "Tourette"), alpha = FDR)
save_deg_results(res_Bx3, "M3.3_Basal_OCD_vs_TS")

## Global LRT — Basal: any diagnosis differs?

dds_basal_lrt <- DESeqDataSetFromMatrix(countData = counts_basal, colData = coldata_basal, design = ~ diagnosis)

dds_basal_lrt <- DESeq(dds_basal_lrt, test = "LRT", reduced = ~ 1)

res_lrt_basal <- results(dds_basal_lrt, alpha = FDR)
save_deg_results_lrt(res_lrt_basal, "M3_LRT_Basal_global")

```

```{r}
# Comparar modelo sin batch vs con batch en M3
dds_basal_batch <- DESeqDataSetFromMatrix(
  countData = counts_basal,
  colData   = coldata_basal,
  design    = ~ batch + diagnosis
)
dds_basal_batch <- DESeq(dds_basal_batch)

res_batch <- results(dds_basal_batch,
                     contrast = c("diagnosis", "TEA", "Tourette"),
                     alpha    = FDR)

# Cuántos genes significativos con batch en el modelo
sig_batch <- sum(res_batch$padj < FDR &
                 abs(res_batch$log2FoldChange) > FC,
                 na.rm = TRUE)
message("Genes sig con batch: ", sig_batch)
message("Genes sig sin batch: ", 98)

# Ver si los top genes anteriores siguen siendo significativos
res_batch_df <- as.data.frame(res_batch) %>%
  rownames_to_column("gene") %>%
  filter(gene %in% top5)

print(res_batch_df %>% select(gene, log2FoldChange, padj))
```
## 8.2 Diagnosis differences Basal + shrinkage and bacth effect

~ batch + diagnosis in Basal subset
Comparisons global + pairwise (ASD-OCD, ASD-TS, OCD-TS) without LPS noise

```{r diagnosis basal}

idx_basal     <- coldata$treatment == "Basal"
counts_basal  <- counts[, idx_basal]
coldata_basal <- coldata[idx_basal, ]
coldata_basal$diagnosis <- droplevels(coldata_basal$diagnosis)
coldata_basal$batch     <- droplevels(coldata_basal$batch)

dds_basal <- DESeqDataSetFromMatrix(countData = counts_basal, colData = coldata_basal, design = ~ batch + diagnosis)
dds_basal <- DESeq(dds_basal)

message("Basal model coefficients:")
print(resultsNames(dds_basal))

# ── Función de guardado con ashr ──────────────────────────────
save_deg_results_ashr <- function(dds, contrast_vec, name) {

  res <- results(dds, contrast = contrast_vec, alpha = FDR)
  res_shrunk <- lfcShrink(dds, contrast = contrast_vec, type = "ashr", res = res)
  
  res_df <- as.data.frame(res_shrunk) %>%
    rownames_to_column("gene") %>%
    arrange(padj)

  res_df$status <- "NS"
  res_df$status[res_df$padj < FDR & res_df$log2FoldChange >  FC] <- "Up"
  res_df$status[res_df$padj < FDR & res_df$log2FoldChange < -FC] <- "Down"
  res_df$status <- factor(res_df$status, levels = c("Up", "Down", "NS"))

  n_up   <- sum(res_df$status == "Up",   na.rm = TRUE)
  n_down <- sum(res_df$status == "Down", na.rm = TRUE)
  message(name, " — Up: ", n_up, "  Down: ", n_down)

  write.csv(res_df, file = file.path(out_dir, paste0(name, "_results.csv")), row.names = FALSE)

  res_filt <- res_df %>% filter(!is.na(padj))

  # Volcano
  p_vol <- ggplot(res_filt,
                  aes(x = log2FoldChange, y = -log10(padj),
                      color = status)) +
    geom_point(size = 1.2, alpha = 0.6) +
    scale_color_manual(values = c("Up"   = "#e05c5c",
                                   "Down" = "#4e9af1",
                                   "NS"   = "grey70")) +
    geom_vline(xintercept = c(-FC, FC),
               linetype = "dashed", color = "grey40") +
    geom_hline(yintercept = -log10(FDR),
               linetype = "dashed", color = "grey40") +
    annotate("text",
             x = max(res_filt$log2FoldChange) * 0.8,
             y = max(-log10(res_filt$padj), na.rm = TRUE) * 0.95,
             label = paste0("Up: ", n_up),
             color = "#e05c5c", size = 4) +
    annotate("text",
             x = min(res_filt$log2FoldChange) * 0.8,
             y = max(-log10(res_filt$padj), na.rm = TRUE) * 0.95,
             label = paste0("Down: ", n_down),
             color = "#4e9af1", size = 4) +
    labs(title    = paste("Volcano —", name),
         subtitle = paste0("FDR < ", FDR,
                           "  |  |log2FC| > ", FC,
                           "  |  ashr shrinkage"),
         x = "log2 Fold Change (shrunk)", y = "-log10(FDR)") +
    theme_bw(base_size = 13) +
    theme(plot.title     = element_text(face = "bold"),
          legend.title   = element_blank())

  ggsave(file.path(plot_dir, paste0(name, "_volcano.png")),
         p_vol, width = 7, height = 6, dpi = 300)

  # MA plot
  p_ma <- ggplot(res_filt,
                 aes(x = log2(baseMean + 1), y = log2FoldChange,
                     color = status)) +
    geom_point(size = 1, alpha = 0.5) +
    scale_color_manual(values = c("Up"   = "#e05c5c",
                                   "Down" = "#4e9af1",
                                   "NS"   = "grey70")) +
    geom_hline(yintercept = c(-FC, 0, FC),
               linetype = c("dashed", "solid", "dashed"),
               color = "grey40") +
    labs(title    = paste("MA plot —", name),
         subtitle = paste0("FDR < ", FDR,
                           "  |  |log2FC| > ", FC,
                           "  |  ashr shrinkage"),
         x = "log2(mean expression)", y = "log2 Fold Change (shrunk)") +
    theme_bw(base_size = 13) +
    theme(plot.title   = element_text(face = "bold"),
          legend.title = element_blank())

  ggsave(file.path(plot_dir, paste0(name, "_MA.png")),
         p_ma, width = 7, height = 6, dpi = 300)

  invisible(res_df)
}

# ── Comparaciones M3 ──────────────────────────────────────────

save_deg_results_ashr(dds_basal, c("diagnosis", "TEA", "TOC"), "M3.1_Basal_ASD_vs_OCD")

save_deg_results_ashr(dds_basal, c("diagnosis", "TEA", "Tourette"), "M3.2_Basal_ASD_vs_TS")

save_deg_results_ashr(dds_basal, c("diagnosis", "TOC", "Tourette"), "M3.3_Basal_OCD_vs_TS")

```


## 9. Diagnosis differences LPS

~ diagnosis in LPS subset
Comparisons global + pairwise (ASD-OCD, ASD-TS, OCD-TS)

```{r diagnosis}

idx_lps     <- coldata$treatment == "LPS"
counts_lps  <- counts[, idx_lps]
coldata_lps <- coldata[idx_lps, ]
coldata_lps$diagnosis <- droplevels(coldata_lps$diagnosis)

dds_lps <- DESeqDataSetFromMatrix(countData = counts_lps, colData = coldata_lps, design = ~ diagnosis)
dds_lps <- DESeq(dds_lps)

# ASD vs OCD

res_L1 <- results(dds_lps, contrast = c("diagnosis", "TEA", "TOC"), alpha = FDR)
save_deg_results(res_L1, "M4.1_LPS_ASD_vs_OCD")

# ASD vs TS

res_L2 <- results(dds_lps, contrast = c("diagnosis", "TEA", "Tourette"), alpha = FDR)
save_deg_results(res_L2, "M4.2_LPS_ASD_vs_TS")

# OCD vs TS

res_L3 <- results(dds_lps, contrast = c("diagnosis", "TOC", "Tourette"), alpha = FDR)
save_deg_results(res_L3, "M4.3_LPS_OCD_vs_TS")

## Global LRT — Basal: any diagnosis differs?

dds_lps_lrt <- DESeqDataSetFromMatrix(countData = counts_lps, colData = coldata_lps, design = ~ diagnosis)

dds_lps_lrt <- DESeq(dds_lps_lrt, test = "LRT", reduced = ~ 1)

res_lrt_lps <- results(dds_lps_lrt, alpha = FDR)
save_deg_results_lrt(res_lrt_lps, "M4_LRT_LPS_global")
```

# Integration LRT results

```{r}
sig_basal <- rownames(res_lrt_basal)[!is.na(res_lrt_basal$padj) & res_lrt_basal$padj < FDR]
sig_lps   <- rownames(res_lrt_lps)[!is.na(res_lrt_lps$padj)   & res_lrt_lps$padj   < FDR]

venn_list <- list("Basal LRT" = sig_basal, "LPS LRT" = sig_lps)

p_venn <- ggVennDiagram(venn_list, label_alpha = 0, set_size = 5) +
  scale_fill_gradient(low = "#f7fbff", high = "#b03060") + coord_flip() + 
  labs(title = "Overlap between diagnosis LRT (M3 / M4)") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

ggsave(file.path(plot_dir, "M6_ANOVA_venn_LRT.png"),
       p_venn, width = 9, height = 6, dpi = 300)
```

## 9.2 Diagnosis differences LPS

~ btach + diagnosis in LPS subset
Comparisons global + pairwise (ASD-OCD, ASD-TS, OCD-TS)

```{r diagnosis lps}

idx_lps     <- coldata$treatment == "LPS"
counts_lps  <- counts[, idx_lps]
coldata_lps <- coldata[idx_lps, ]
coldata_lps$diagnosis <- droplevels(coldata_lps$diagnosis)
coldata_lps$batch     <- droplevels(coldata_lps$batch)

dds_lpsb <- DESeqDataSetFromMatrix(countData = counts_lps, colData = coldata_lps, design = ~ batch + diagnosis)
dds_lpsb <- DESeq(dds_lpsb)

message("LPS model coefficients:")
print(resultsNames(dds_lpsb))

# ── Comparaciones M4 ──────────────────────────────────────────

save_deg_results_ashr(dds_lpsb, c("diagnosis", "TEA", "TOC"), "M4.4_LPS_ASD_vs_OCD")

save_deg_results_ashr(dds_lpsb, c("diagnosis", "TEA", "Tourette"), "M4.5_LPS_ASD_vs_TS")

save_deg_results_ashr(dds_lpsb, c("diagnosis", "TOC", "Tourette"), "M4.6_LPS_OCD_vs_TS")
```

```{r}
# ── Comparar los tres modelos en M4 TEA vs TOC como ejemplo ──

# Modelo original sin batch ni shrinkage
res_original <- results(dds_lps,  # tu dds_lps sin batch
                        contrast = c("diagnosis", "TEA", "TOC"),
                        alpha    = FDR)

# Modelo con batch sin shrinkage
res_batch <- results(dds_lpsb,
                     contrast = c("diagnosis", "TEA", "TOC"),
                     alpha    = FDR)

# Modelo con batch + ashr
res_ashr <- lfcShrink(dds_lpsb,
                      contrast = c("diagnosis", "TEA", "TOC"),
                      type     = "ashr",
                      res      = res_batch)

# Contar significativos en cada modelo
count_sig <- function(res, label) {
  n <- sum(res$padj < FDR & abs(res$log2FoldChange) > FC,
           na.rm = TRUE)
  message(label, ": ", n, " genes")
}

count_sig(res_original, "Sin batch, sin shrinkage")
count_sig(res_batch,    "Con batch, sin shrinkage")
count_sig(res_ashr,     "Con batch, con ashr")

# ── Ver distribución de LFC antes y después de shrinkage ──────
plot_df <- data.frame(
  gene         = rownames(res_batch),
  LFC_original = res_batch$log2FoldChange,
  LFC_ashr     = res_ashr$log2FoldChange,
  padj         = res_ashr$padj
) %>%
  filter(!is.na(LFC_original), !is.na(LFC_ashr))

# Scatter LFC original vs LFC ashr
ggplot(plot_df, aes(x = LFC_original, y = LFC_ashr,
                    color = !is.na(padj) & padj < FDR)) +
  geom_point(size = 0.8, alpha = 0.4) +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", color = "grey40") +
  scale_color_manual(values = c("FALSE" = "grey70", "TRUE" = "#e05c5c"),
                     labels = c("FALSE" = "ns", "TRUE" = "sig")) +
  labs(title    = "M4 TEA vs TOC — efecto del shrinkage ashr",
       subtitle = "Puntos por encima de la diagonal = LFC inflado antes de shrinkage",
       x = "LFC sin shrinkage", y = "LFC con ashr",
       color = NULL) +
  theme_bw(base_size = 12)

# ── Comparar LFC de genes significativos antes vs después ─────
sig_genes <- plot_df %>%
  filter(!is.na(padj), padj < FDR, abs(LFC_ashr) > FC) %>%
  arrange(padj)

message("\nTop 10 genes significativos con ashr:")
print(sig_genes %>%
        select(gene, LFC_original, LFC_ashr, padj) %>%
        slice_head(n = 10))

# ── Ver cuántos genes pierden significancia solo por FC < 1 ───
# Es decir, padj < FDR pero LFC shrunk < 1
lost_by_fc <- plot_df %>%
  filter(!is.na(padj), padj < FDR,
         abs(LFC_original) > FC,
         abs(LFC_ashr) < FC)

message("\nGenes que pierden significancia solo porque",
        " ashr contrae LFC por debajo de 1: ",
        nrow(lost_by_fc))
print(lost_by_fc %>%
        select(gene, LFC_original, LFC_ashr, padj) %>%
        arrange(desc(abs(LFC_original))))
```

```{r}
# Modificar save_deg_results_ashr para M3/M4
# Quitar el filtro de FC — ashr ya controla los LFC inflados

save_deg_results_ashr <- function(dds, contrast_vec, name) {

  res <- results(dds,
                 contrast = contrast_vec,
                 alpha    = FDR)

  res_shrunk <- lfcShrink(dds,
                           contrast = contrast_vec,
                           type     = "ashr",
                           res      = res)

  res_df <- as.data.frame(res_shrunk) %>%
    rownames_to_column("gene") %>%
    arrange(padj)

  # Sin filtro de FC — ashr ya controla inflación
  res_df$status <- "NS"
  res_df$status[res_df$padj < FDR & res_df$log2FoldChange >  0] <- "Up"
  res_df$status[res_df$padj < FDR & res_df$log2FoldChange <  0] <- "Down"
  res_df$status <- factor(res_df$status, levels = c("Up", "Down", "NS"))

  n_up   <- sum(res_df$status == "Up",   na.rm = TRUE)
  n_down <- sum(res_df$status == "Down", na.rm = TRUE)
  message(name, " — Up: ", n_up, "  Down: ", n_down)

  write.csv(res_df,
            file      = file.path(out_dir, paste0(name, "_results.csv")),
            row.names = FALSE)

  res_filt <- res_df %>% filter(!is.na(padj))

  # Volcano — sin líneas de FC
  p_vol <- ggplot(res_filt,
                  aes(x = log2FoldChange, y = -log10(padj),
                      color = status)) +
    geom_point(size = 1.2, alpha = 0.6) +
    scale_color_manual(values = c("Up"   = "#e05c5c",
                                   "Down" = "#4e9af1",
                                   "NS"   = "grey70")) +
    geom_hline(yintercept = -log10(FDR),
               linetype = "dashed", color = "grey40") +
    annotate("text",
             x = max(res_filt$log2FoldChange, na.rm = TRUE) * 0.8,
             y = max(-log10(res_filt$padj),   na.rm = TRUE) * 0.95,
             label = paste0("Up: ", n_up),
             color = "#e05c5c", size = 4) +
    annotate("text",
             x = min(res_filt$log2FoldChange, na.rm = TRUE) * 0.8,
             y = max(-log10(res_filt$padj),   na.rm = TRUE) * 0.95,
             label = paste0("Down: ", n_down),
             color = "#4e9af1", size = 4) +
    labs(title    = paste("Volcano —", name),
         subtitle = paste0("FDR < ", FDR,
                           "  |  ashr shrinkage  |  no FC threshold"),
         x = "log2 Fold Change (shrunk)", y = "-log10(FDR)") +
    theme_bw(base_size = 13) +
    theme(plot.title   = element_text(face = "bold"),
          legend.title = element_blank())

  ggsave(file.path(plot_dir, paste0(name, "_volcano.png")),
         p_vol, width = 7, height = 6, dpi = 300)

  # MA plot
  p_ma <- ggplot(res_filt,
                 aes(x = log2(baseMean + 1), y = log2FoldChange,
                     color = status)) +
    geom_point(size = 1, alpha = 0.5) +
    scale_color_manual(values = c("Up"   = "#e05c5c",
                                   "Down" = "#4e9af1",
                                   "NS"   = "grey70")) +
    geom_hline(yintercept = 0, color = "grey40") +
    labs(title    = paste("MA plot —", name),
         subtitle = paste0("FDR < ", FDR,
                           "  |  ashr shrinkage  |  no FC threshold"),
         x = "log2(mean expression)",
         y = "log2 Fold Change (shrunk)") +
    theme_bw(base_size = 13) +
    theme(plot.title   = element_text(face = "bold"),
          legend.title = element_blank())

  ggsave(file.path(plot_dir, paste0(name, "_MA.png")),
         p_ma, width = 7, height = 6, dpi = 300)

  invisible(res_df)
}
```



## 10. Comparison LPS change between samples

logFC (Basal -> LPS) per donor

```{r donor-LPS}

# TMM normalization before logCPM
dge_norm <- DGEList(counts = counts)
dge_norm <- calcNormFactors(dge_norm, method = "TMM")
logcpm   <- cpm(dge_norm, log = TRUE, prior.count = 1)

# Apply per-donor logFC (Basal -> LPS)

donors <- unique(coldata$donor_id)

lfc_list <- lapply(donors, function(d) {
  s_basal <- rownames(coldata)[coldata$donor_id == d & coldata$treatment == "Basal"]
  s_lps   <- rownames(coldata)[coldata$donor_id == d & coldata$treatment == "LPS"]
  
  # solo eliminar si falta completamente una condición
  if (length(s_basal) == 0 || length(s_lps) == 0) return(NULL)
  
  basal_mean <- rowMeans(logcpm[, s_basal, drop = FALSE])
  lps_mean   <- rowMeans(logcpm[, s_lps, drop = FALSE])
  
  lps_mean - basal_mean
})

names(lfc_list) <- donors
lfc_list <- lfc_list[!sapply(lfc_list, is.null)]

# seguridad
if (length(lfc_list) == 0) {
  stop("No valid donors found after filtering")
}

lfc_matrix <- do.call(cbind, lfc_list)
colnames(lfc_matrix) <- names(lfc_list)

# Donor-level coldata (one row per donor)

coldata_donor <- coldata[match(colnames(lfc_matrix), as.character(coldata$donor_id)), ]
rownames(coldata_donor) <- coldata_donor$donor_id
coldata_donor$diagnosis <- droplevels(coldata_donor$diagnosis)

message("Per-donor LFC matrix: ", nrow(lfc_matrix), " genes x ",
        ncol(lfc_matrix), " donors")
print(table(coldata_donor$diagnosis))

# limma
design_lfc <- model.matrix(~ 0 + diagnosis, data = coldata_donor)
colnames(design_lfc) <- levels(coldata_donor$diagnosis)

fit_lfc  <- lmFit(lfc_matrix, design_lfc)
cm_lfc   <- makeContrasts(
  TEA_vs_TOC      = TEA - TOC,
  TEA_vs_Tourette = TEA - Tourette,
  TOC_vs_Tourette = TOC - Tourette,
  levels = design_lfc
)
fit_lfc2 <- contrasts.fit(fit_lfc, cm_lfc)
fit_lfc2 <- eBayes(fit_lfc2)

# Save each contrast
for (ct in colnames(cm_lfc)) {
  res_lfc <- topTable(fit_lfc2, coef = ct, number = Inf, sort.by = "P") %>%
    rownames_to_column("gene") %>%
    rename(log2FoldChange = logFC,
           padj           = adj.P.Val,
           pvalue         = P.Value,
           baseMean       = AveExpr)

  res_lfc$status <- "NS"
  res_lfc$status[res_lfc$padj < FDR & res_lfc$log2FoldChange >  FC] <- "Up"
  res_lfc$status[res_lfc$padj < FDR & res_lfc$log2FoldChange < -FC] <- "Down"

  n_up   <- sum(res_lfc$status == "Up")
  n_down <- sum(res_lfc$status == "Down")
  cname  <- paste0("M5_LFC_response_", ct)
  message(cname, " — Up: ", n_up, "  Down: ", n_down)

  write.csv(res_lfc,
            file = file.path(out_dir, paste0(cname, "_results.csv")),
            row.names = FALSE)

  # Volcano
  p_vol <- ggplot(res_lfc, aes(x=log2FoldChange, y=-log10(padj), color=status)) +
    geom_point(size=1.2, alpha=0.6) +
    scale_color_manual(values=c("Up"="#e05c5c","Down"="#4e9af1","NS"="grey70")) +
    geom_vline(xintercept=c(-FC, FC), linetype="dashed", color="grey40") +
    geom_hline(yintercept=-log10(FDR),    linetype="dashed", color="grey40") +
    labs(title    = paste("Volcano —", cname),
         subtitle = paste0("FDR < ", FDR, "  |  |log2FC| > ", FC),
         x="log2 Fold Change", y="-log10(FDR)") +
    theme_bw(base_size=13) +
    theme(plot.title=element_text(face="bold"), legend.title=element_blank())

  ggsave(file.path(plot_dir, paste0(cname, "_volcano.png")),
         p_vol, width=7, height=6, dpi=300)
}

```


```{r}

# TMM normalization before logCPM
dge_norm <- DGEList(counts = counts)
dge_norm <- calcNormFactors(dge_norm, method = "TMM")
logcpm   <- cpm(dge_norm, log = TRUE, prior.count = 1)

# Apply per-donor logFC (Basal -> LPS)

donors <- unique(coldata$donor_id)

lfc_list <- lapply(donors, function(d) {
  s_basal <- rownames(coldata)[coldata$donor_id == d & coldata$treatment == "Basal"]
  s_lps   <- rownames(coldata)[coldata$donor_id == d & coldata$treatment == "LPS"]
  
  # solo eliminar si falta completamente una condición
  if (length(s_basal) == 0 || length(s_lps) == 0) return(NULL)
  
  basal_mean <- rowMeans(logcpm[, s_basal, drop = FALSE])
  lps_mean   <- rowMeans(logcpm[, s_lps, drop = FALSE])
  
  lps_mean - basal_mean
})

names(lfc_list) <- donors
lfc_list <- lfc_list[!sapply(lfc_list, is.null)]

# seguridad
if (length(lfc_list) == 0) {
  stop("No valid donors found after filtering")
}

lfc_matrix <- do.call(cbind, lfc_list)
colnames(lfc_matrix) <- names(lfc_list)

# Donor-level coldata (one row per donor)

coldata_donor <- coldata[match(colnames(lfc_matrix), as.character(coldata$donor_id)), ]
rownames(coldata_donor) <- coldata_donor$donor_id
coldata_donor$diagnosis <- droplevels(coldata_donor$diagnosis)

message("Per-donor LFC matrix: ", nrow(lfc_matrix), " genes x ",
        ncol(lfc_matrix), " donors")
print(table(coldata_donor$diagnosis))

# limma
fit_lfc  <- lmFit(lfc_matrix, design_lfc)

fit_lfc_F <- eBayes(fit_lfc)

topTable(fit_lfc_F, number = Inf, sort.by = "F")

anova_results <- topTable(fit_lfc_F, number = Inf, sort.by = "F") %>%
  rownames_to_column("gene") %>%
  rename(
    baseMean = AveExpr,
    pvalue   = P.Value,
    padj     = adj.P.Val
  ) %>%
  mutate(
    status = ifelse(padj < FDR, "Significant", "NS")
  )

# Guardar CSV
write.csv(
  anova_results,
  file = file.path(out_dir, "M5.2_ANOVA_all_genes.csv"),
  row.names = FALSE
)
message("Tabla completa de ANOVA guardada: ANOVA_all_genes.csv")

# ==============================
# 5️⃣ Gráfico global tipo volcano-like
# ==============================
ggplot(anova_results, aes(x=baseMean, y=-log10(padj), color=status)) +
  geom_point(alpha=0.6, size=1.5) +
  scale_color_manual(values=c("Significant"="#e05c5c","NS"="grey70")) +
  geom_hline(yintercept=-log10(FDR), linetype="dashed", color="grey40") +
  labs(
    title="ANOVA global — todos los genes",
    x="Average expression (logCPM)",
    y="-log10(FDR)"
  ) +
  theme_bw(base_size=13) +
  theme(legend.title = element_blank(), plot.title = element_text(face="bold"))


```

```{r}


# =============================
# 0️⃣ Normalización TMM y logCPM
# ==============================
dge_norm <- DGEList(counts = counts)
dge_norm <- calcNormFactors(dge_norm, method = "TMM")
logcpm <- cpm(dge_norm, log = TRUE, prior.count = 1)

# ==============================
# 1️⃣ Calcular logFC por donante (Basal -> LPS)
# ==============================
donors <- unique(coldata$donor_id)

lfc_list <- lapply(donors, function(d) {
  s_basal <- rownames(coldata)[coldata$donor_id == d & coldata$treatment == "Basal"]
  s_lps <- rownames(coldata)[coldata$donor_id == d & coldata$treatment == "LPS"]

  if (length(s_basal) == 0 || length(s_lps) == 0) return(NULL)
  rowMeans(logcpm[, s_lps, drop=FALSE]) - rowMeans(logcpm[, s_basal, drop=FALSE])
})

names(lfc_list) <- donors
lfc_list <- lfc_list[!sapply(lfc_list, is.null)]
lfc_matrix <- do.call(cbind, lfc_list)
colnames(lfc_matrix) <- names(lfc_list)

coldata_donor <- coldata[match(colnames(lfc_matrix), as.character(coldata$donor_id)), ]
rownames(coldata_donor) <- coldata_donor$donor_id
coldata_donor$diagnosis <- droplevels(coldata_donor$diagnosis)

message("Per-donor LFC matrix: ", nrow(lfc_matrix), " genes x ", ncol(lfc_matrix), " donors")
print(table(coldata_donor$diagnosis))

# ==============================
# 2️⃣ ANOVA global
# ==============================
design_lfc <- model.matrix(~ 0 + diagnosis, data = coldata_donor)
fit_lfc <- lmFit(lfc_matrix, design_lfc)
fit_lfc_F <- eBayes(fit_lfc)

anova_results <- topTable(fit_lfc_F, number = Inf, sort.by = "F") %>%
  rownames_to_column("gene") %>%
  rename(baseMean = AveExpr, pvalue = P.Value, padj = adj.P.Val) %>%
  mutate(status = ifelse(padj < FDR, "Significant", "NS"))

write.csv(anova_results, file.path(out_dir, "M5.3.ANOVA_all_genes.csv"), row.names = FALSE)
message("ANOVA global guardada: M5.3.ANOVA_all_genes.csv")

# ==============================
# 3️⃣ Filtrar genes significativos para contrastes
# ==============================
sig_genes <- anova_results$gene[anova_results$padj < FDR]
lfc_matrix_sig <- lfc_matrix[sig_genes, ]

# ==============================
# 4️⃣ Contrasts
# ==============================
fit_lfc2 <- lmFit(lfc_matrix_sig, design_lfc)
cm_lfc <- makeContrasts(
  TEA_vs_TOC      = diagnosisTEA - diagnosisTOC,
  TEA_vs_Tourette = diagnosisTEA - diagnosisTourette,
  TOC_vs_Tourette = diagnosisTOC - diagnosisTourette,
  levels = design_lfc
)
fit_lfc2 <- contrasts.fit(fit_lfc2, cm_lfc)
fit_lfc2 <- eBayes(fit_lfc2)

# ==============================
# 5️⃣ Guardar resultados y volcano plots
# ==============================
for (ct in colnames(cm_lfc)) {
  res_lfc <- topTable(fit_lfc2, coef = ct, number = Inf, sort.by = "P") %>%
    rownames_to_column("gene") %>%
    rename(
      log2FoldChange = logFC,
      padj = adj.P.Val,
      pvalue = P.Value,
      baseMean = AveExpr
    )

  res_lfc$status <- "NS"
  res_lfc$status[res_lfc$padj < FDR & res_lfc$log2FoldChange > FC] <- "Up"
  res_lfc$status[res_lfc$padj < FDR & res_lfc$log2FoldChange < -FC] <- "Down"

  n_up <- sum(res_lfc$status == "Up")
  n_down <- sum(res_lfc$status == "Down")
  cname <- paste0("M5.3_LFC_response_", ct)
  message(cname, " — Up: ", n_up, "  Down: ", n_down)

  # Guardar CSV
  write.csv(res_lfc, file.path(out_dir, paste0(cname, "_results.csv")), row.names = FALSE)

  # Volcano plot
  p_vol <- ggplot(res_lfc, aes(x = log2FoldChange, y = -log10(padj), color = status)) +
    geom_point(size=1.2, alpha=0.6) +
    scale_color_manual(values = c("Up"="#e05c5c", "Down"="#4e9af1", "NS"="grey70")) +
    geom_vline(xintercept=c(-FC, FC), linetype="dashed", color="grey40") +
    geom_hline(yintercept=-log10(FDR), linetype="dashed", color="grey40") +
    labs(title = paste("Volcano —", cname),
         subtitle = paste0("FDR < ", FDR, " | |log2FC| > ", FC),
         x = "log2 Fold Change", y = "-log10(FDR)") +
    theme_bw(base_size=13) +
    theme(plot.title = element_text(face="bold"), legend.title = element_blank())

  ggsave(file.path(plot_dir, paste0(cname, "_volcano.png")), p_vol, width = 7, height = 6, dpi = 300)
}

```



